# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
1. Read in the activity data.
2. Convert the "date" column into the Date class.
```{r pre-process, echo=TRUE, results='hide'}
mydata <- read.csv("~/R/Coursera/Reproducible-Research/data/repdata-data-activity/activity.csv", stringsAsFactors = F)
mydata$date <- as.Date(mydata$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
Tally up the steps taken each day.
```{r}
step_sum <- aggregate(steps ~ date, data = mydata, sum)
```

Create a histogram of the total steps per day.
```{r histogram, fig.width=7, fig.height=6}
hist(step_sum$steps, breaks=10)
```

The mean number of total steps per day is:
```{r}
step_sum_mean <- mean(step_sum$steps)
print(step_sum_mean)
```

The median number of total steps per day is:
```{r}
step_sum_median <- median(step_sum$steps)
print(step_sum_median)
```




## What is the average daily activity pattern?
Average the number of steps for each 5 minute interval across all days.
```{r}
interval_steps <- aggregate(steps ~ interval, data = mydata, mean)
```

Plot the data.
```{r lineplot, fig.width=7, fig.height=6}
plot(interval_steps$interval, interval_steps$steps, type = "l")
```

The interval with the highest number of steps on average is:
```{r}
interval_steps[interval_steps$steps==max(interval_steps$steps),]
```


## Imputing missing values
The total number of missing values (missing steps per day/interval) is:
```{r}
sum(is.na(mydata$steps))
```

Impute all missing values by replacing "NA" with the mean number of steps for that specific 5 minute interval (mean across all days).
```{r}
# renaming a column due to merging two datasets, each with a column named "steps".   
colnames(interval_steps)[2] <- "mean.int.steps"
impdata <- merge(mydata, interval_steps, by="interval")
impdata <- impdata[order(impdata$date,impdata$interval),]
head(impdata)

impdata$steps[is.na(impdata$steps)] <- impdata$mean.int.steps
head(impdata)
```

Tally up the steps taken each day (with imputed values).
```{r}
impstep_sum <- aggregate(steps ~ date, data = impdata, sum)
```

Create a histogram of the total steps per day (with imputed values).
```{r histogram imputed, fig.width=7, fig.height=6}
hist(impstep_sum$steps, breaks=10)
```

The mean number of total steps per day (with imputed values):
```{r}
impstep_sum_mean <- mean(impstep_sum$steps)
print(impstep_sum_mean)
```

The median number of total steps per day is (with imputed values):
```{r}
impstep_sum_median <- median(impstep_sum$steps)
print(impstep_sum_median)
```


Imputing missing values changed the mean number of steps per day by:
```{r}
difmean <- impstep_sum_mean - step_sum_mean
print(difmean)
```

Imputing missing values changed the median number of steps per day by:
```{r}
difmedian <- impstep_sum_median - step_sum_median
print(difmedian)
```

## Are there differences in activity patterns between weekdays and weekends?

Add a weekday column to the data with imputed missing values.
```{r}
impdata$wkday <- weekdays(impdata$date)

# build a separate data.frame containing days of the week
wkday <- c("Sunday", "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")
daytype <- c("weekend","weekday","weekday","weekday","weekday","weekday","weekend")
wk <- data.frame(wkday, daytype)
print(wk)

# now merge the wkday data.frame with the original data with imputed missing values.
impdata_wk <- merge(impdata, wk, by = "wkday", sort = F)
impdata_wk <- impdata_wk[order(impdata_wk$date, impdata_wk$interval),]

```


Now average for each interval by weekend
```{r}
imp_interval_steps <- aggregate(steps ~ interval + daytype, data = impdata_wk, mean)
```

Plot the data.
```{r panel lineplot, fig.width=7, fig.height=6}
library(lattice)
xyplot(steps ~ interval | daytype, data = imp_interval_steps, type="l", layout=c(1,2))
```